Edge sign prediction in general graphs and Correlation Clustering

### 2017-12-02 cc/gtx_algorithm.tex
tiny fixes and ref all over the places
review gtx description and experiments
in particular, avoid mentioning strech by name, and say we only have a
conjecture regarding the runtime
also make the transition between cc and gtx smoother

To achieve our objective of building a spanning tree that minimizes the distances between all
connected pair of nodes in the original graph, we rely on a particular subgraph structure, namely
the star.  We therefore introduce two algorithmic primitives: \extractStar{}, which partition a
graph $G$ into a set of disjoint stars; and \collapseStar{}, which selects edges from $E$ to
assemble these stars into a new, smaller graph. Given a graph topology $G_0=(V_0, E_0)$ and assuming
for simplicity that $G_0$ consists of a single connected component,\footnote{For we can otherwise
run our algorithm in parallel on each connected components of $G_0$.} the \gtx{} algorithm
repeatedly applies these two primitives to produce a sequence of graphs $\{G_t\}_{t=0}^K$ of
decreasing size, until $G_K$ is made of a single node. All the edges selected while reaching
this stage then form the spanning tree we were looking for.
This can be seen as a simple instantiation of a general procedure described in \autocite[Section
5.2]{LowerBound95}.
% [from a CMU NIPS submission http://www.stat.cmu.edu/~arinaldo/papers/mutualfriends.pdf] We
% consider Mutual-Friends to be *an algorithmic primitive*, by which we mean a kind of subroutine for
% a more complicated function that iterates Mutual-Friends, similarly to [5].

In the following, we provide a more precise description of our two primitives and analyze their
complexity. Then we state formally the complete \gtx{} algorithm, prove its termination and
correctness, and show a detailed example of its execution.  Finally, we study its properties, such
as the number of iterations needed to finish and the stretch of the resulting tree.

@@

As showed in the following pseudo code\footnote{Note that for
clarity, we removed some bookkeeping code in all listings, mainly the part related to maintaining
mapping between nodes at different collapse level. However, the full python implementation
is available at \url{https://github.com/daureg/magnet/blob/master/veverica/new_galaxy.py\#L27}.}, we
sample centers by choosing the node with the current highest degree, with ties broken
arbitrarily.\footnote{We also consider more involved heuristics but, in the interest of simplicity,
they are presented later~\vpageref{ssec:gtx_center_choice}.} This is achieved efficiently by
maintaining a max-priority queue $Q$, initially containing all the nodes of $V_t$. The priority of a
node is its current degree and we equip $Q$ with two standard operations described
by~\textcite[section 6.5]{CormenAlgo09}: \textsc{Extract-Max}$(Q)$ removes and returns the
node of $Q$ with the largest degree and \textsc{Decrease-Key}$(Q$, $u$, $\Delta)$ decrements
the degree of the node $u$ by an amount $\Delta$.
We also assume that $G$ is the adjacency list of the graph, so that $G[u]$ is the set of neighbors of
$u$, \ie{} $G[u] \equiv \mathcal{N}(u)$. Finally $membership$ is a map and we
define the \textsc{Star} function, which creates a star given a center $c$, a list $periphery$ of
peripheral nodes, and a star index $i$. After creating the \ith{} star, for every node $u$ belonging
to that star, the \textsc{Star} function sets $membership[u] = i$.
\vspace{-\baselineskip}

@@

\begin{prop}
  \label{prop:gtx_time_extractstar}
  For any connected graph $G=(V,E)$, \extractStar$(G)$ terminates in $O(|E|)$ time.
\end{prop}
\begin{proof}
\extractStar{} terminates because at each iteration of the while loop line 7, we remove one node
from $Q$ and never add any. Let us now analyze its complexity. We first build a priority queue of
all the nodes according to their degree (line 2), which requires $|V|$ insertions into $Q$. Then
we execute $|V|$ iterations of the while loop. However, the main idea here is that we process each
node and each edge exactly once. We first find the center $c$ of the next star by extracting the
maximum of the queue (line 8) and testing if $c$ is still part of the graph, which happens $|V|$
time. Then we build the corresponding star (line 11--14). In total, we test the membership of $|V|$
nodes in line 11, $\textsc{Star}$ updates the $membership$ map $|V|$ times in line 12,
$inner\_edges$ consists of $|E|$ edges at most in line 13 and $remaining$ is only updated $|V|$
times in line 14. Finally, we decrease the priority (\ie{} the degree) of all nodes adjacent to the
new star (line 15--17). Each decrement is supported by a single edge, thus \textsc{Decrease-Keys} is
called at most $|E|$ times. Since all queue operations require constant time when using a Strict
Fibonacci Heap~\autocite{FibonacciHeaps12}, the complexity is $O(|E|+|V|)$, which is also $O(|E|)$
since $G$ is connected.
\end{proof}

@@

The second primitive, \collapseStar{} takes as input the edges $E_t$ of the current graph, along
with the $membership$ result of \extractStar{} and an $eccentricity$ array. It builds a new graph
$G_{t+1}$ where each star from \extractStar{} becomes a node and where there is a link between two
nodes $s_u$ and $s_v$ if the nodes in $V_t$ making up $s_u$ and $s_v$ are connected in $E_t$. The
$eccentricity$ array is needed because when connecting two stars, we would prefer to join their
center rather than two of their peripheral points. Therefore, we maintain an eccentricity count for
all of the nodes of the original $G_0$, which is incremented by $1$ each time a node is chosen to be
on the periphery of a star. In other words, the eccentricity of an original node quantifies to which
extent it has been pushed to the border of the galaxy.

\begin{marginfigure}
  \centering
  \includegraphics[width=0.92\linewidth]{assets/tikz/gtx_cross_edges_tikz.pdf}
  \caption[Cross edge representation]{In this graph, there are four possible edges between the two
  stars $s_1$ and $s_2$, and all are part of $cross\_edges[(s_1, s_2)]$, here represented in light
red. Those edges are labeled with the sum of eccentricity of their underlying endpoints, and the
minimal one is linking a peripheral node of $s_1$ directly to the center of $s_2$.}
  \label{fig:gtx_cross_edge}
\end{marginfigure}%
\collapseStar{} not only returns the new graph $G_{t+1}$ but also a map $cross\_edges$. This map is
used to keep track of which edge in $E_t$ connects any pair of node in $V_{t+1}$ (as illustrated in
\autoref{fig:gtx_cross_edge}). It associates to any edge in $E_{t+1}$ a set of edges from $E_t$.
This set can have an arbitrary size during the execution of \collapseStar{}, yet it is guaranteed to
contain a single edge upon termination. Although this does not appear in the following pseudo code,
in practice we ensure that $(s_u, s_v)$ and $(s_v, s_u)$ refer to the same edge in $cross\_edges$.

\begin{center}
  \rule{\textwidth}{.3pt}
  \begin{algorithmic}[1]
    \Function{\collapseStar{}}{$E_t,\,membership,\,eccentricity$}
      \State Let $G_{t+1}$ be an empty graph
      \State Let $cross\_edges$ be the map described above
      \ForAll{edge $(u, v)$ in $E_t$}
      \Let{$s_u$}{$membership[u]$}
      \Let{$s_v$}{$membership[v]$}
        \If{$u$ and $v$ are not in the same star (\ie{} $s_u \neq s_v$)}
          \Let{$cross\_edges[(s_u, s_v)]$}{$cross\_edges[(s_u, s_v)] \bigcup \{(u, v)\}$}
        \EndIf
      \EndFor
      \ForAll{pair of will-be connected stars $(s_u,s_v)$ in $cross\_edges$}
        \Let{$possible\_underlying\_edges$}{$cross\_edges[(s_u, s_v)]$}
        \marginpars{\small Line 11 is actually a simplification since the $eccentricity$ array is only
        defined for nodes in $V_0$ so when \collapseStar{} is called on $G_3$ for instance, $u$ and $v$
        are nodes of $V_3$ and we have to retrieve the actual corresponding edge in $E_0$. Maybe I could
        simply add a map from $E_t$ to $E_0$ like in the real python code…}
        \State $$(u_0, v_0) \gets \argmin_{(u,v)\in possible\_underlying\_edges} eccentricity[u]+eccentricity[v]$$
        \Let{$cross\_edges[(s_u, s_v)]$}{$\{(u_0, v_0)\}$}
        \Let{$E_{t+1}$}{$E_{t+1} \bigcup \{(s_u, s_v)\}$}
      \EndFor
      \State \textbf{return} $G_{t+1}$, $cross\_edges$
    \EndFunction
  \end{algorithmic}
  \rule{\textwidth}{.3pt}
\end{center}

@@

\paragraph{Putting the pieces together}\label{par:full_gtx}%
\extractStar{} and \collapseStar{} are truly the core the of \gtx{} algorithm but to obtain our
final spanning tree, we need additional work, namely updating the eccentricity of the nodes of
$V_0$ and keeping track of each edge within and between stars along every collapse steps. Despite
our earlier promise, this entails showing some bookkeeping code, because it has an influence on the
runtime of \gtx{}. \Todo{Those maps could probably use better names}
Namely in the listing of \autoref{alg:gtx} \vpageref{alg:gtx}, we use the three following maps:

\vspace{0.5\baselineskip}
\noindent\begin{tabulary}{187mm}{lllL}
  \toprule
  name               & keys set at $t$ & values set at $t$   & comment \\
  \midrule
  $star\_membership$ & nodes in $V_t$  & nodes in $V_{t+1}$  & this is the map returned by
  \extractStar{}$(G_t)$ \\
  $full\_membership$ & nodes in $V_0$  & nodes in $V_{t+1}$  & at $t=0$, this is equals to
  $star\_membership$ and then it gets updated at every iteration to maintain its original keys set.  \\
  $original\_node$   & nodes in $V_t$  & sets of $V_0$ nodes & This can be seen as the reverse of
  $full\_membership$ and it is the one needed to update the eccentricity of the original nodes. \\
  \bottomrule
\end{tabulary}
\vspace{0.5\baselineskip}

@@

As described in \autoref{alg:gtx}, at every collapse level, we first extract stars from the current
graph (line 5), then update the eccentricity and node mappings (lines 7--8) and finally collapse the
graph (line 9). We perform these operations until there are no edge connecting stars anymore. At
this point, we revisit every outer edges to build the spanning tree (line 13).

@@

\begin{proof}[Proof of \autoref{prop:gtx_correct}]
  Let us first show that \gtx{} terminates in less than $|V_0|$ iterations. This follows from the
  fact that every time we collapse the graph $G_t$, we strictly reduce the number of nodes. Indeed,
  according to \autoref{lem:gtx_stay_connected}, $G_t$ is connected so at least two nodes of $V_t$
  are joined by an edge and will form a star, \ie{} a single node. We can thus claim that $|V_{t+1}| <
  |V_t|$. Note also that for all $t$, $|V_t| > 0$ and that when $|V_t|=1$, \extractStar{} creates a
  singleton and \collapseStar{} does not return any outer edges so \gtx{} finishes, proving that the
  number of iterations $K$ satisfies $K \leq |V_0|$.

 Then we analyze the time complexity. We already now that during the \tth{} iteration,
 \extractStar{} and \collapseStar{} take time $O(|E_t|)$. We only provide the actual python code
 instead of pseudo code for the remaining functions since they do not present any relevant
 algorithmic aspect. However, we can say that \textsc{Update-Eccentricity} and\marginpars{We can get
 rid of this $O(|V_0|)$ at each iteration by ignoring eccentricity and taking a random edge between
 star, although it's difficult to tell how it would affect the quality of the resulting tree.}
 \textsc{Update-Nodes-Mapping} take $O(|V_0|)$ time since they go through every node of the original
 graph. Because $G_0$ is connected, $|V_0| \leq |E_0|$, and since for all $t$, $|E_t| \leq |E_0|$,
 we have that the \gtx{} inner loop takes $O(K|E_0|)$ time. Finally, since
 \textsc{Assemble-Spanning-Tree} goes again through every edge visited at every collapse step, it
 also takes $O(K|E_0|)$ time, which is thus the overall complexity of the \gtx{} algorithm.

@@

The bound of \autoref{prop:gtx_correct} amounts to $O(|V_0||E_0|)$, which we believe is overly
pessimistic. First because we expect the number of nodes to decrease by more than one at each step,
maybe leading to only $O(\log |V_0|)$ iterations. More importantly, instead of bounding
$\sum_{t=0}^K |E_t|$ by $K|E_0|$, we hope to show that $E_{t+1} = rE_t$ with $r<1$ so that this sum
is $O(|E_0|)$.

@@

We now run \extractStar{} on $G_1$. Because all nodes have degree $2$, they could all be chosen to
be the center of a star yet we again they are picked according to their index and therefore we
choose $\starone{1}$ to be the center of the star $\startwo{1}$ with peripheral nodes $\starone{2}$
and $\starone{3}$. The original nodes belonging to those peripheral stars (nodes $4$ to $15$) have
their eccentricity incremented by $1$. The next node with highest degree in $G_1$ is now
$\starone{4}$, which forms a star with $\starone{5}$ and $\starone{6}$. This choice means that nodes
$21$ through $30$ have their eccentricity incremented by $1$. Finally, $\starone{7}$ forms the last
star with $\starone{8}$ and $\starone{9}$. Then \collapseStar{} connects the resulting three stars,
and this time there is only a single choice between each pair of stars, leading to the graph $G_2$
showed in \autoref{fig:gtx_run2}

The action of \extractStar{} on $G_2$ is quite simple, because there is only one star that can be
created, so let say we choose $\startwo{1}$ as its center, with $\startwo{2}$ and $\startwo{3}$ as
peripheral nodes. This increases the eccentricity of their underlying $G_0$ nodes by $1$ (namely
nodes $16$ to $45$). Because there is only one star $\starthree{1}$ left, \collapseStar{} returns
$G_3$ showed in \autoref{fig:gtx_run3} and an empty list of outer edges, meaning that the inner loop
of \gtx{} is finished and we can go through every edges we chose between stars at every level to
recover the final spanning tree, a process illustrated in \autoref{fig:gtx_spanning_tree}
\vpageref{fig:gtx_spanning_tree}. For completeness, we can also look at the edges which are not part
of the spanning tree and therefore contribute to the stretch of the tree. In that case the average
stretch is $7$, as showed in \autoref{tab:gtx_example_stretch}.

\begin{table}[htpb]
  \centering
  \caption{Stretch of the example tree}
  \label{tab:gtx_example_stretch}
  \begin{tabulary}{\linewidth}{lLr}
    \toprule
    test edge & path in the tree & length \\
    \midrule
    $2,9$   & $2$--$6$--$9$                       & $2$ \\
    $3,12$  & $3$--$1$--$4$--$14$--$11$--$12$     & $6$ \\
    $13,29$ & $13$--$11$--$15$--$26$--$29$        & $4$ \\
    $20,24$ & $20$--$16$--$17$--$23$--$21$--$24$  & $5$ \\
    $25,42$ & $25$--$21$--$23$--$17$--$16$--$19$--$29$--$26$--$15$--$11$--$14$--$4$--$1$--$2$--$6$--$8$--$37$--$36$--$39$-- $32$--$31$--$35$--$43$--$41$--$45$ & $24$ \\
    $33,38$ & $33$--$31$--$32$--$39$--$36$--$38$  & $5$ \\
    $34,43$ & $34$--$31$--$35$--$43$              & $3$ \\
    \bottomrule
  \end{tabulary}
\end{table}

\paragraph{Number of iterations needed}\label{par:number_of_iteration}%

\textcolor{red}{\LARGE Draft}

A crucial quantity of the \gtx{} algorithm, both in terms of complexity and resulting stretch, is
the number of collapses $K$ needed before termination. While this is still elusive to express in the
general case, let us first look at some simple cases. For instance, a very sparse example of graph
is the line graph. While it is already a tree, let us look how the \gtx{} algorithm operates on it.
Say we have $n$ nodes and $m$ edges in that line. Here, a star is made at most of three
consecutive nodes. In the worst case, centers will be chosen such that we have a succession of three
and two nodes stars, as in \autoref{fig:gtx_line_graph}.%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.9\linewidth]{assets/tikz/gtx_line_tikz.pdf}
  \caption{A line graph with stars in blue}
  \label{fig:gtx_line_graph}
\end{marginfigure}
This will results in $\nicefrac{2n}{5}$ stars, which is less than half of $n$ and because in that
case $n=m+1$, \gtx{} will finish after $O(\log m)$ iterations. Note that a barbell graph (two
cliques connected by a line) would requires a number of iterations logarithmic in the length of that
central line, despite having many more edges. This suggests unsurprisingly that the diameter of the graph
could be a good parameter to quantify the number of iterations needed. Indeed, take a tree and consider
the length $p$ of its longest path from the root to a leaf. By a similar argument as the one used in
the line case, it seems \gtx{} will terminate after $O(\log p)$ iterations.

@@

\begin{figure*}[htpb]
   \centering
   \includegraphics[width=\linewidth]{raw/expand_path.pdf}
   \caption[Star expansion of a path]{A path within $G_t$ in black, and the underlying graph
     $G_{t-1}$ in blue. In the worst case, the endpoints of the path in $G_{t-1}$ are at the
     periphery of the endpoint stars.  The intermediary nodes in $G_t$ could be proper or singleton
     star, but there cannot be two adjacent singleton stars.}
   \label{fig:gtx_expand_path}
\end{figure*}

@@

A variant of the \gtx{} algorithm as a whole we did not explore so much in practice is its ability
to produce spanner by stopping early. Basically if we stop at iteration $t$, we output the graph
$G_t$ (which in general is not a tree) with its edges unfolded to lie in $E_0$. This corresponds to
a trade off between having more edges than $|V_0|-1$ but potentially making shorter connection and
thus having a lower stretch. This would also be interesting in the sign prediction setting. Assuming
the treewidth of $G_t$ is decreasing with $t$, maybe we could show there is \enquote{reasonable}
number of paths between $u$ and $v$, compute their parity and take a majority vote on the results.

>>>>

To achieve our objective of building a spanning tree that minimizes the distances between all
connected pair of nodes in the original graph, we rely on a particular subgraph structure, namely
the star.  We therefore introduce two algorithmic primitives. \extractStar{}, which partition a
graph $G$ into a set of disjoint stars. And \collapseStar{}, which selects edges from $E$ to
assemble these stars into a new, smaller graph. Given a graph topology $G_0=(V_0, E_0)$ and assuming
for simplicity that $G_0$ consists of a single connected component,\footnote{For we can otherwise
run our algorithm in parallel on each connected components of $G_0$.} the \gtx{} algorithm
repeatedly applies these two primitives to produce a sequence of graphs $\{G_t\}_{t=0}^K$ of
decreasing size, until $G_K$ is made of a single node. All the edges selected while reaching
this stage then form the spanning tree we were looking for.
This can be seen as a simple instantiation of a general procedure described in \autocite[Section
5.2]{LowerBound95}.
% [from a CMU NIPS submission http://www.stat.cmu.edu/~arinaldo/papers/mutualfriends.pdf] We
% consider Mutual-Friends to be *an algorithmic primitive*, by which we mean a kind of subroutine for
% a more complicated function that iterates Mutual-Friends, similarly to [5].

In the following, we provide a more precise description of our two primitives and analyze their
complexity. Then we state formally the complete \gtx{} algorithm, prove its termination and
correctness, and show a detailed example of its execution.  Finally, we provide a conjecture on
the number of iterations needed to finish.

@@

As showed in the following pseudo code\footnote{Note that for
clarity, we removed some bookkeeping code in all listings, mainly the part related to maintaining
mapping between nodes at different collapse level. However, the full python implementation
is available at \url{https://github.com/daureg/magnet/blob/master/veverica/new_galaxy.py\#L27}.}, we
sample centers by choosing the node with the current highest degree, with ties broken
arbitrarily.\footnote{We also consider more involved heuristics but, in the interest of simplicity,
they are presented later~\vpageref{ssec:gtx_center_choice}.} This is achieved efficiently by
maintaining a max-priority queue $Q$, initially containing all the nodes of $V_t$. The priority of a
node is its current degree and we equip $Q$ with two standard operations described
by~\textcite[section 6.5]{CormenAlgo09}: \textsc{Extract-Max}$(Q)$ removes and returns the
node of $Q$ with the largest degree and \textsc{Decrease-Key}$(Q$, $u$, $\Delta)$ decrements
the degree of the node $u$ by an amount $\Delta$.
We also assume that $G$ is the adjacency list of the graph, so that $G[u]$ is the set of neighbors of
$u$, \ie{} $G[u] \equiv \mathcal{N}(u)$. Finally $membership$ is the map storing for each node the
index of the star it belongs to. This map is updated by the
\textsc{Star} function, which creates a star given a center $c$, a list $periphery$ of
peripheral nodes, and a star index $i$. After creating the \ith{} star, for every node $u$ belonging
to that star, the \textsc{Star} function sets $membership[u] = i$.
\vspace{-\baselineskip}

@@

\begin{prop}
  \label{prop:gtx_time_extractstar}
  For any connected graph $G=(V,E)$, \extractStar$(G)$ terminates in $O(|E|)$ time.
\end{prop}
\begin{proof}
\extractStar{} terminates because at each iteration of the while loop line 7, we remove one node
from $Q$ and never add any. Let us now analyze its complexity. We first build a priority queue of
all the nodes according to their degree (line 2), which requires $|V|$ insertions into $Q$. Then
we execute $|V|$ iterations of the while loop. However, the main idea here is that we process each
node and each edge exactly once. We first find the center $c$ of the next star by extracting the
maximum of the queue (line 8) and testing if $c$ is still part of the graph, which happens $|V|$
times in total. Then we build the corresponding star (line 11--14). In total, we test the membership
of $|V|$
nodes in line 11, $\textsc{Star}$ updates the $membership$ map $|V|$ times in line 12,
$inner\_edges$ consists of $|E|$ edges at most in line 13 and $remaining$ is only updated $|V|$
times in line 14. Finally, we decrease the priority (\ie{} the degree) of all nodes adjacent to the
new star (line 15--17). Each decrement is supported by a single edge, thus \textsc{Decrease-Keys} is
called at most $|E|$ times. Since all queue operations require constant time when using a Strict
Fibonacci Heap~\autocite{FibonacciHeaps12}, the complexity is $O(|E|+|V|)$, which is also $O(|E|)$
since $G$ is connected.
\end{proof}

@@

The second primitive, \collapseStar{} takes as input the edges $E_t$ of the current graph, along
with the $membership$ result of \extractStar{} and an $eccentricity$ array. It builds a new graph
$G_{t+1}$ where each star $s_u$ from \extractStar{} becomes a node and where there is a link between two
nodes $s_u$ and $s_v$ if the nodes in $V_t$ making up $s_u$ and $s_v$ are connected in $E_t$. The
$eccentricity$ array is needed because when connecting two stars, we would prefer to join their
center rather than two of their peripheral points. Therefore, we maintain an eccentricity count for
all of the nodes of the original $G_0$, which is incremented by $1$ each time a node is chosen to be
on the periphery of a star. In other words, the eccentricity of an original node quantifies to which
extent it has been pushed to the border of the galaxy.

\begin{marginfigure}
  \centering
  \includegraphics[width=0.92\linewidth]{assets/tikz/gtx_cross_edges_tikz.pdf}
  \caption[Cross edge representation]{In this graph, there are four possible edges between the two
  stars $s_1$ and $s_2$, and all are part of $cross\_edges[(s_1, s_2)]$, here represented in light
red. Those edges are labeled with the sum of eccentricity of their underlying endpoints. White node
have an eccentricity of $0$, and gray node an eccentricity of $1$. The edge with the
minimal eccentricity is linking a peripheral node of $s_1$ directly to the center of $s_2$.}
  \label{fig:gtx_cross_edge}
\end{marginfigure}%
\collapseStar{} not only returns the new graph $G_{t+1}$ but also a map $cross\_edges$. This map is
used to keep track of which edge in $E_t$ connects any pair of node in $V_{t+1}$ (as illustrated in
\autoref{fig:gtx_cross_edge}). It associates to any edge in $E_{t+1}$ a set of edges from $E_t$.
This set can have an arbitrary size during the execution of \collapseStar{}, yet it is guaranteed to
contain a single edge upon termination. Although this does not appear in the following pseudo code,
in practice we ensure that $(s_u, s_v)$ and $(s_v, s_u)$ refer to the same edge in $cross\_edges$.

\begin{center}
  \rule{\textwidth}{.3pt}
  \begin{algorithmic}[1]
    \Function{\collapseStar{}}{$E_t,\,membership,\,eccentricity$}
      \State Let $G_{t+1}$ be an empty graph
      \State Let $cross\_edges$ be the map described above
      \For{\textbf{every} edge $(u, v)$ in $E_t$}
      \Let{$s_u$}{$membership[u]$}
      \Let{$s_v$}{$membership[v]$}
        \If{$u$ and $v$ are not in the same star (\ie{} $s_u \neq s_v$)}
          \Let{$cross\_edges[(s_u, s_v)]$}{$cross\_edges[(s_u, s_v)] \bigcup \{(u, v)\}$}
        \EndIf
      \EndFor
      \For{\textbf{every} pair of will-be connected stars $(s_u,s_v)$ in $cross\_edges$}
        \Let{$possible\_underlying\_edges$}{$cross\_edges[(s_u, s_v)]$}
        %\marginpars{\small Line 11 is actually a simplification since the $eccentricity$ array is
        %only defined for nodes in $V_0$ so when \collapseStar{} is called on \eg{} $G_3$, $u$
        %and $v$ are nodes of $V_3$ and we have to retrieve the actual corresponding edge in $E_0$.
        %Maybe I could simply add a map from $E_t$ to $E_0$ like in the real python code…}
        \State $$(u_0, v_0) \gets \argmin_{(u,v)\in possible\_underlying\_edges} eccentricity[u]+eccentricity[v]$$
        \Let{$cross\_edges[(s_u, s_v)]$}{$\{(u_0, v_0)\}$}
        \Let{$E_{t+1}$}{$E_{t+1} \bigcup \{(s_u, s_v)\}$}
      \EndFor
      \State \textbf{return} $G_{t+1}$, $cross\_edges$
    \EndFunction
  \end{algorithmic}
  \rule{\textwidth}{.3pt}
\end{center}

@@

\paragraph{Putting the pieces together}\label{par:full_gtx}%
\extractStar{} and \collapseStar{} are truly the core the of \gtx{} algorithm but to obtain our
final spanning tree, we need additional work, namely updating the eccentricity of the nodes of
$V_0$ and keeping track of each edge within and between stars along every collapse steps. Despite
our earlier promise, this entails showing some bookkeeping code, because it has an influence on the
runtime of \gtx{}. %\Todo{Those maps could probably use better names}
Namely in the listing of \autoref{alg:gtx} \vpageref{alg:gtx}, we use the three following maps:

\vspace{0.5\baselineskip}
\setlength{\fullpage}{\textwidth+\marginparsep+\marginparwidth}
\noindent\begin{tabulary}{\fullpage}{lllL}
  \toprule
  name               & keys set at $t$ & values set at $t$   & comment \\
  \midrule
  $star\_membership$ & nodes in $V_t$  & nodes in $V_{t+1}$  & this is the map returned by
  \extractStar{}$(G_t)$ \\
  $full\_membership$ & nodes in $V_0$  & nodes in $V_{t+1}$  & at $t=0$, this is equals to
  $star\_membership$ and then it gets updated at every iteration to maintain its original keys set.  \\
  $original\_node$   & nodes in $V_t$  & sets of $V_0$ nodes & This can be seen as the reverse of
  $full\_membership$ and it is the one needed to update the eccentricity of the original nodes. \\
  \bottomrule
\end{tabulary}
\vspace{0.5\baselineskip}

@@

As described in \autoref{alg:gtx}, at every collapse level, we first extract stars from the current
graph (line 5), then update the eccentricity and node mappings (lines 7--8) and finally collapse the
graph (line 9). We perform these operations until there are no edge connecting stars anymore. At
this point, we revisit every outer edge to build the spanning tree (line 13).

@@

\begin{proof}[Proof of \autoref{prop:gtx_correct}]
  Let us first show that \gtx{} terminates in less than $|V_0|$ iterations. This follows from the
  fact that every time we collapse the graph $G_t$, we strictly reduce the number of nodes. Indeed,
  according to \autoref{lem:gtx_stay_connected}, $G_t$ is connected so at least two nodes of $V_t$
  are joined by an edge and will form a star, \ie{} a single node. We can thus claim that $|V_{t+1}| <
  |V_t|$. Note also that for all $t$, $|V_t| > 0$ and that when $|V_t|=1$, \extractStar{} creates a
  singleton and \collapseStar{} does not return any outer edges. At this point, \gtx{} finishes,
  proving that the number of iterations $K$ satisfies $K \leq |V_0|$.

 Then we analyze the time complexity. We already now that during the \tth{} iteration,
 \extractStar{} and \collapseStar{} take time $O(|E_t|)$. We only provide the actual python code
 instead of pseudo code for the remaining functions since they do not present any relevant
 algorithmic aspect. However, we can say that \textsc{Update-Eccentricity} and\marginpars{We can get
 rid of this $O(|V_0|)$ at each iteration by ignoring eccentricity and taking a random edge between
 star, although it's difficult to tell how it would affect the quality of the resulting tree.}
 \textsc{Update-Nodes-Mapping} take $O(|V_0|)$ time, since they go through every node of the original
 graph. Because $G_0$ is connected, $|V_0| \leq |E_0|$, and since for all $t$, $|E_t| \leq |E_0|$,
 we have that the \gtx{} inner loop takes $O(K|E_0|)$ time. Finally, since
 \textsc{Assemble-Spanning-Tree} goes again through every edge visited at every collapse step, it
 also takes $O(K|E_0|)$ time, which is thus the overall complexity of the \gtx{} algorithm.

@@

The bound of \autoref{prop:gtx_correct} amounts to $O(|V_0||E_0|)$, which we believe is overly
pessimistic. Indeed, we expect the number of nodes to decrease by more than one at each step,
potentially leading to only $O(\log |V_0|)$ iterations.

@@

We now run \extractStar{} on $G_1$. Because all nodes have degree $2$, they could all be chosen to
be the center of a star. We again assume they are picked according to their index and therefore we
choose $\starone{1}$ to be the center of the star $\startwo{1}$ with peripheral nodes $\starone{2}$
and $\starone{3}$. The original nodes belonging to those peripheral stars (nodes $4$ to $15$) have
their eccentricity incremented by $1$. The next node with highest degree in $G_1$ is now
$\starone{4}$, which forms a star with $\starone{5}$ and $\starone{6}$. This choice means that nodes
$21$ through $30$ have their eccentricity incremented by $1$. Finally, $\starone{7}$ forms the last
star with $\starone{8}$ and $\starone{9}$. Then \collapseStar{} connects the resulting three stars,
and this time there is only a single choice between each pair of stars, leading to the graph $G_2$
showed in \autoref{fig:gtx_run2}

The action of \extractStar{} on $G_2$ is quite simple, because there is only one star that can be
created, so let say we choose $\startwo{1}$ as its center, with $\startwo{2}$ and $\startwo{3}$ as
peripheral nodes. This increases the eccentricity of their underlying $G_0$ nodes by $1$ (namely
nodes $16$ to $45$). Because there is only one star $\starthree{1}$ left, \collapseStar{} returns
$G_3$ showed in \autoref{fig:gtx_run3} and an empty list of outer edges, meaning that the inner loop
of \gtx{} is finished and we can go through every edges we chose between stars at every level to
recover the final spanning tree, a process illustrated in \autoref{fig:gtx_spanning_tree}
\vpageref{fig:gtx_spanning_tree}. For completeness, we can also look at the edges which are not part
of the spanning tree, that is the test edges, and compute how long are the paths joining their
endpoints. Indeed, this is the second quantity appearing in the mistakes bound of
equation~\eqref{eq:stretch_mistakes}. As shown in \autoref{tab:gtx_example_stretch}, this average
length is $7$.

\begin{table}[htpb]
  \centering
  \caption{Length of the paths not in the resulting spanning tree.}
  \label{tab:gtx_example_stretch}
  \begin{tabulary}{\linewidth}{lLr}
    \toprule
    test edge & path in the tree & length \\
    \midrule
    $2,9$   & $2$--$6$--$9$                       & $2$ \\
    $3,12$  & $3$--$1$--$4$--$14$--$11$--$12$     & $6$ \\
    $13,29$ & $13$--$11$--$15$--$26$--$29$        & $4$ \\
    $20,24$ & $20$--$16$--$17$--$23$--$21$--$24$  & $5$ \\
    $25,42$ & $25$--$21$--$23$--$17$--$16$--$19$--$29$--$26$--$15$--$11$--$14$--$4$--$1$--$2$--$6$--$8$--$37$--$36$--$39$-- $32$--$31$--$35$--$43$--$41$--$45$ & $24$ \\
    $33,38$ & $33$--$31$--$32$--$39$--$36$--$38$  & $5$ \\
    $34,43$ & $34$--$31$--$35$--$43$              & $3$ \\
    \bottomrule
  \end{tabulary}
\end{table}

\paragraph{Discussion on the number of iterations needed}\label{par:number_of_iteration}%

A crucial quantity of the \gtx{} algorithm, both in terms of complexity and resulting stretch, is
the number of collapses $K$ needed before termination. While this is still elusive to express in the
general case, let us first look at some simple cases. For instance, a very sparse example of graph
is the line graph. While it is already a tree, let us look how the \gtx{} algorithm operates on it.
Say we have $n$ nodes and $m$ edges in that line. Here, a star is made at most of three
consecutive nodes. In the worst case, centers will be chosen such that we have a succession of three
and two nodes stars, as in \autoref{fig:gtx_line_graph}.%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.9\linewidth]{assets/tikz/gtx_line_tikz.pdf}
  \caption{A line graph with stars in blue.}
  \label{fig:gtx_line_graph}
\end{marginfigure}
This will results in $\nicefrac{2n}{5}$ stars organized in a line, which is less than half of $n$.
Thus we can conclude that in this case, \gtx{} will finish after $O(\log n)$ iterations. Note that a barbell graph (two
cliques connected by a line) would requires a number of iterations logarithmic in the length of that
central line, despite having many more nodes and edges.

In fact, it is helpful to look at this question from the other direction. Namely, we can wonder what
is the minimum number $n(K)$ of nodes necessary for a graph to requires at least $K$ collapses
before being reduced to a single node. Checking by hand, we have that $n(1)=2$, $n(2)=4$ and
$n(3)=8$. Furthermore, this number is achieved by a line or circle consisting of exactly $2^K$
nodes. By induction, we can even show that on any line or circle graphs with $2^n$ nodes, the \gtx{}
algorithm will terminate after $n$ collapses. Assuming we could prove that $n(K)=2^K$, given a graph
$G$ with $n=2^{\log_2 n}$ nodes, by definition of $n(K)=2^K$, it would require less than
$\ceil{\log_2 n} = O(\log n)$ collapses before being reduced to a single node. However, we were not
able to rule out the possibility that there is a graph more parsimonious than a line still able to
survive the same number of collapses. Therefore, we cannot conclude that $n(K)=2^K$ and we leave
this statement as a conjecture.

\iffalse
This suggests unsurprisingly that the diameter of the graph
could be a good parameter to quantify the number of iterations needed. Indeed, take a tree and consider
the length $p$ of its longest path from the root to a leaf. By a similar argument as the one used in
the line case, it seems \gtx{} will terminate after $O(\log p)$ iterations.

@@

\begin{figure*}[htpb]
   \centering
   \includegraphics[width=\linewidth]{raw/expand_path.pdf}
   \caption[Star expansion of a path]{A path within $G_t$ in black, and the underlying graph
     $G_{t-1}$ in blue. In the worst case, the endpoints of the path in $G_{t-1}$ are at the
     periphery of the endpoint stars.  The intermediary nodes in $G_t$ could be proper or singleton
     star, but there cannot be two adjacent singleton stars.}
   \label{fig:gtx_expand_path}
\end{figure*}
\fi

@@

A variant of the \gtx{} algorithm as a whole we did not explore so much in practice is its ability
to produce spanners by stopping early. Basically if we stop at iteration $t$, we output the graph
$G_t$ (which in general is not a tree) with its edges unfolded to lie in $E_0$. This corresponds to
a trade off between having more edges than $|V_0|-1$ but potentially making shorter connection.
This would also be interesting \esp{} problem. Assuming
the treewidth of $G_t$ is decreasing with $t$, we could try to show there is \enquote{reasonable}
number of paths between $u$ and $v$, compute their parity and take a majority vote on the results.
